---
title: "Project 1"
author: "Maxim Arisov"
date: "2026-09-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Project 1

Approach

For this project, I will use the chess tournament data provided for the assignment. My goal is to clean the data and create a CSV file with information about each player.

First, I will read the tournament data into R and look at how the data is set and organized.

Next, I will find each player's name, state, total points, pre tournament rating, and opponents.

Then, I will use the opponent numbers to find each opponent's pre tournament rating. I will calculate the average opponent rating for each player. For example, Gary Hua's average opponent rating is 1605.

Finally, I will create a clean table with the following items:

Player Name, State, Total Points, Pre Rating, and Average Opponent Pre Rating.

At last, I will check the results and save the final table as a CSV file.

Code Base

```{r}
install.packages("pdftools")
library(pdftools)
```

```{r}
#Lets show PDF location on gitHub.
url <- "https://raw.githubusercontent.com/Maximus2485/Project-1-.DATA-607/main/Project%201%20dataset.tournamentinfo.pdf"
```

```{r}
#Lets download PDF file
download.file(url, "Project 1 dataset.tournamentinfo.pdf", mode="wb")
# Lets read PDF
chess <- pdf_text("Project 1 dataset.tournamentinfo.pdf")
head(chess)
```

```{r}
#Combine PDF pages
chess_text <- paste(chess, collapse="\n")
# Lets split the dataset into lines
chess_lines <- unlist(strsplit(chess_text, "\n"))
head(chess_lines, 20)
```

```{r}
#Lets find player information in the lines having "I" and total number with a 
# decimal point
player_lines <- chess_lines[
  grepl("\\|", chess_lines) &
  grepl("[0-9]+\\.[0-9]", chess_lines)
]

# Check player lines
head(player_lines)
```

```{r}
#Lets find the rating/state lines
rating_lines <- chess_lines[grep("R:", chess_lines)]
head(rating_lines)
```

```{r}
#Lets extract players names
player_name <- trimws(
  sub("\\|.*", "",
  sub("^\\ s*[0-9]+\\|", "", player_lines))
)
player_name
```

```{r}
#Lets extract total points
points <- sapply(strsplit(player_lines, "\\|"), function(x) {
  as.numeric(trimws(x[3]))
})

points
```

```{r}
#Lets extract state
state <- trimws(
  sub("\\|.*", "", rating_lines)
)
state
```

```{r}
#Lets extract pre-rating
pre_rating <- as.numeric(
  sub(".*R:\\s*([0-9]+).*", "\\1", rating_lines)
)
pre_rating
```

```{r}
#Lets create the basic player table
players <- data.frame(
  Player = player_name,
  State = state,
  Points = points,
  Pre_Rating = pre_rating
)
head(players)
```

```{r}
#Lets extract opponent numbers
get_opponents <- function(x) {
  
  parts <- strsplit(x, "\\|")[[1]]
  
  rounds <- parts[4:length(parts)]
  
  opponents <- as.numeric(gsub("[^0-9]","", rounds)
                          
                          )
  
  opponents[!is.na(opponents)]
}
#Lets apply it to every player:
opponents <- lapply(player_lines, get_opponents)
opponents[[1]]

```

```{r}
#Lets calculate average opponent pre-rating
average_opponent_rating <- sapply(opponents, function(x) {
  mean(players$Pre_Rating[x],
       na.rm=TRUE )
})
#Add it to the table
players$Avg_Opponent_Rating <- round(average_opponent_rating)

#Check the final results
head(players)
```

```{r}
#Lets use SQLite
install.packages("DBI")
install.packages("RSQLite")
library(DBI)
library(RSQLite)

# Lets create database connection
con <- dbConnect(SQLite(), "Project 1 dataset.tournamentinfo.db")

# Lets put players data into SQL database
dbWriteTable(con, "players", players, overwrite = TRUE)
```


SQL Query
```{sql connection=con}
SELECT
    Player,
    State,
    Points,
    Pre_Rating,
    Avg_Opponent_Rating
FROM players;

```


```{r}
#Lets export SQL results to CSV file
final_data <- dbGetQuery(
  con,"SELECT
    Player,
    State,
    Points,
    Pre_Rating,
    Avg_Opponent_Rating
FROM players"
)
write.csv(final_data, "chess_results.csv", row.names = FALSE)
```

```{r}
dbDisconnect(con)
```

References

Google Deep Mind. (2026). Gemini 3 Flash [Large language model]. [[https://gemini.google.com.%5D(https://gemini.google.com/](https://gemini.google.com.%5D(https://gemini.google.com/)%7B.uri%7D) Accessed September 27th, 2026.

Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2023). R for data science (2nd ed.). O'Reilly Media. <https://r4ds.hadley.nz/>
